Proof nets for first-order additive linear logic

### 2019-05-31 7 - FSCD 2019 - Final/ALL1.tex
trying to match editor's version

\documentclass[UKenglish]{lipics-v2019}

@@

\usepackage{amsmath,amssymb,amsthm}
\usepackage{latexsym}
\usepackage{microtype}
\usepackage{proof}
\usepackage{refcount}
\usepackage{stmaryrd}
\usepackage{xargs}

@@

\newcommand\defn[1]{\textit{\textbf{#1}}}

@@

\newcommand\gen{\leq}
\newcommand\coh{\smallfrown}
\newcommand\join{\vee}
\newcommand\minus{\mathop{\!/\mathchoice{\kern-3pt}{\kern-3pt}{\kern-2.5pt}{\kern-2pt}/\!}}

@@

\author
 {Willem B.\ Heijltjes}
 {University of Bath, United Kingdom}
 {{http://willem.heijltj.es}} %{w.b.heijltjes@bath.ac.uk}
 {}
 {}
\author
 {Dominic J.\ D.\ Hughes}
 {Logic Group, UC Berkeley, USA}
 {{http://boole.stanford.edu/\textasciitilde dominic}} %{dominich@theory.stanford.edu}
 {}
 {} 
\author
 {Lutz Stra\ss burger}
 {Inria Saclay \&\ LIX/\'Ecole Polytechnique, Palaiseau, France}
 {{http://www.lix.polytechnique.fr/Labo/Lutz.Strassburger}}
 %{lutz.strassburger@inria.fr}
 {}
 {}

@@

@@

\paragraph*{Additive proof nets}
%
\all\ proof nets \cite[Section 4.10]{Hughes-vanGlabbeek-2005} represent a morphism $A\to B$ by a sequent $\seq{\dual A}B$ plus a \defn{linking}, a relation between the propositional atoms of $\dual A$ (the dual of $A$) and those of $B$. They are \defn{canonical}: they factor out the permutations of sequent calculus, and correspond 1--1 to morphisms of the free category with binary sums and products. \defn{Composition}, of proof nets over $\seq{\dual A}B$ and $\seq{\dual B}C$ to one over $\seq{\dual A}C$, is by the relational composition of their linkings along the dual formulas $B$ and $\dual B$ and captures sequent-calculus cut-elimination. Below are examples of proof nets and their composition.
%
\begin{center}\vspace{-.5ex}\(
\begin{array}{@{}rc@{\qquad}rc@{}}
	\text{\small diagonal:}
&	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{{\dual a}}
		\formula[y=0]{a\*a}
		\Vlink[red]{[-2]1,1}
		\Vlink[red]{[2]1,3}
	\end{tikzpicture}}
&	\text{\small injection:}	
&   \vc{\begin{tikzpicture}[net]
		\formula[y=1]{{\dual a}}
		\formula[y=0]{a\+b}
		\Vlink[red]{1,1}
	\end{tikzpicture}}
\\ \\
	\text{\small symmetry:}
&	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{{\dual a}\+{\dual b}}
		\formula[y=0]{b\*a}
		\Vlink[red]{1,3}
		\Vlink[red]{3,1}
	\end{tikzpicture}}
&	\text{\small associativity:}
&	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{*{\phantom(}{\dual a}*{\phantom)}
		 {{\+}}*({\dual b}*{\phantom)}
		 {{\+}}*{\phantom(}{\dual c}*)}
		\formula[y=0]{*(a*{\phantom)}
		{{\*}}*{\phantom(}b*)
		{{\*}}*{\phantom(}c*{\phantom)}}
		\Vlink[red]{1,1}
		\Vlink[red]{3,3}
		\Vlink[red]{5,5}
	\end{tikzpicture}}
\end{array}
\qquad
	\text{\small composition:}
	\quad
    \vc{\begin{tikzpicture}[net]
        \formula[y=2.5]{{\dual b}*\+{\dual a}}
        \formula[y=1.5]{a*\**(b*\+b*)}
            \Vlink[red]{1,2}
            \Vlink[red]{2,1}
        \formula[y=1]{{\dual a}*\+*({\dual b}*\*{\dual b}*)}
        \formula[y=0]{b*\*b}
            \Vlink[blue]{[-1]2,[-1]1}
            \Vlink[blue]{[1]2,[-1]2}
            \Vlink[blue]{[-1]3,[1]1}
            \Vlink[blue]{[1]3,[1]2}
    \end{tikzpicture}}
~~\Rightarrow~~
    \vc{\begin{tikzpicture}[net]
        \formula[y=2.5]{{\dual b}*\+{\dual a}}
        \formula[y=0]{b*\*b}
            \Vlink[violet]{[-1]1,1}
            \Vlink[violet]{[1]1,2}
    \end{tikzpicture}}
\rule[-55pt]{0pt}{110pt}
\)\vspace{-.5ex}\end{center}
%
We extend additive proof nets with first-order quantification. Our central challenge is to incorporate the essential content of first-order proof, the \defn{witness assignment} to existential quantifiers. Commonly, as in the sequent calculus rule below left, a witness to $\exists x.B$ is given by an immediate substitution $B[t/x]$. To assign different witnesses in different branches of a proof, the subformula is duplicated first, giving $B[s/x]$ and $B[t/x]$, as below right.
%
\begin{center}\(
%\hspace*{\fill}
%\hspace*{-\parindent}
	\vc{\infer[\Qrr?t]{\seq A{\exists x.B}}{\seq A{B[t/x]}}}
\hspace{10ex}
\vc{
	\infer[\Qrr*]{\Seq{\exists x.\dual P(x)}{P(s)\*P(t)}}{
	 \infer[\Qrr?s]{\Seq{\exists x.\dual P(x)}{P(s)}}{\infer{\Seq{\dual P(s)}{P(s)}}{}}
	 &&
	 \infer[\Qrr?t]{\Seq{\exists x.\dual P(x)}{P(t)}}{\infer{\Seq{\dual P(t)}{P(t)}}{}}
   }}
%\hfill
\)\end{center}
%
This is incompatible with a \defn{sequent\,+\,links} proof net design, where the conclusion sequent remains intact, and a subformula $B$ cannot be the subject of substitution or duplication. Instead, we propose two alternative treatments of witnessing terms, embodied in two notions of proof net: \defn{witness nets} and \defn{unification nets}. Our solutions are based on the second author's recent \defn{unification nets} for first-order multiplicative linear logic \cite{Hughes-2018}. Their main feature is to omit existential witnesses altogether, and reconstruct them by unification.
%

@@

\defn{Unification nets} omit any witness information, as illustrated below right. In addition to canonicity, they embody a notion of \defn{generality}: where more than one witness could be assigned, unification nets do not require a definite choice, while witness nets do. Composition is direct, by relational composition. We compare further properties in Figure~\ref{fig:results} in the conclusion, where we also discuss related work and Lambek's notion of \defn{generality} \cite{Lambek-1968-1972}.
%
\begin{center}\(
	\text{\small a witness net:}
	\quad
	\vc{\begin{tikzpicture}[net]
		\formula[y=1.2]{\exists x.{\dual P}(x)}
		\formula[y=0]{P(s)\*P(t)}
		\Vlink[red,l,label={$\scriptstyle[s/x]~~$}]{[-3]4,1}
		\Vlink[red,r,label={$~\scriptstyle[t/x]$}]{[1]4,6}
	\end{tikzpicture}}
\hspace{11ex}
	\text{\small a unification net:}
	\quad
	\vc{\begin{tikzpicture}[net]
		\formula[y=1.2]{\exists x.{\dual P}(x)}
		\formula[y=0]{P(s)\*P(t)}
		\Vlink[red]{[-3]4,1}
		\Vlink[red]{[1]4,6}
	\end{tikzpicture}}
\)\vspace{-1ex}\end{center}
%
\paragraph*{Background}
%
Additive linear logic is combinatorially rich, yet well-behaved and tractable: proof search \cite{Galmiche-Marion-1995} and proof net correctness \cite{Heijltjes-Hughes-2015} for a net over $\seq AB$ are linear in $|A|\times|B|$ (with $|A|$ the size of the syntax tree of $A$). Proof nets remain canonical and equally tractable when extended with the two units \cite{Heijltjes-2011,Heijltjes-Hughes-2015}, and the first-order case is merely NP-complete \cite{Heijltjes-Hughes-2015}.

@@

At the heart of a theory of proof nets is the question of \emph{proof identity}: when are two proofs equivalent? The answer determines which proofs should map onto the same proof net. The introduction of quantifiers creates an interesting issue: if two proofs differ by an immaterial choice of existential witness, should they be equivalent? For example, to prove the sequent $\seq{\exists x.P(x)}{\exists y.\dual P(y)}$ both quantifiers must receive \emph{the same} witness, as in the following two proofs, but \emph{any} witness will do.
\begin{center}\(
	\vc{\infer={\seq{\exists x.P(x)}{\exists y.\dual P(y)}}{\infer{\seq{P(s)}{\dual P(s)}}{}}}
	\quad
	\stackrel?\equiv
	\quad
	\vc{\infer={\seq{\exists x.P(x)}{\exists y.\dual P(y)}}{\infer{\seq{P(t)}{\dual P(t)}}{}}}
\)\end{center}
The issue is more pronounced where quantifiers are \emph{vacuous}, $\exists x.A$ with $x$ not free in $A$. The proofs below left can only be distinguished even syntactically because the $\exists\mathrm R$ rule makes the instantiating witness explicit. Below right is an interesting intermediate variant: the witness $s$ or $t$ can be observed without explicit annotation in the $\exists\mathrm R$ rule, but the choice is equally immaterial to the content of the proof as when the quantifier were vacuous.
\begin{center}\(
	\vc{\infer[\!\!\scriptstyle{\qrr?s}]{\seq{\exists x.P}{\dual P}}{\infer{\seq P{\dual P}}{}}}
	\quad
	\stackrel?\equiv
	\quad	
	\vc{\infer[\!\!\scriptstyle{\qrr?t}]{\seq{\exists x.P}{\dual P}}{\infer{\seq P{\dual P}}{}}}\qquad
	\vc{
	 \infer[\!\!\scriptstyle{\qrr?s}]{\seq{\exists x.P\+Q(x)}{\dual P}}{
	  \infer[\!\!\scriptstyle{\qrr+1}]{\seq{P\+Q(s)}{\dual P}}{\infer{\seq P{\dual P}}{}}
	}}
	\quad
	\stackrel?\equiv
	\quad	
	\vc{
	 \infer[\!\!\scriptstyle{\qrr?t}]{\seq{\exists x.P\+Q(x)}{\dual P}}{
	  \infer[\!\!\scriptstyle{\qrr+1}]{\seq{P\+Q(t)}{\dual P}}{\infer{\seq P{\dual P}}{}}
	}}
\)\end{center}
In this paper we will not attempt to settle the question of proof identity. Rather, our two notions of proof net each represent a natural and coherent perspective, at either end of the spectrum. \emph{Witness nets} make all existential witnesses explicit, including those to vacuous quantifiers, rejecting all three equivalences above. \emph{Unification nets} leave all witnesses implicit, thus identifying all proofs modulo witness assignment, and validating all three equivalences.

@@

Additive proof nets have two natural correctness criteria. \defn{Coalescence} \cite{Heijltjes-Hughes-2015,Hughes-Heijltjes-2016}, a counterpart to multiplicative \emph{contractibility} \cite{Danos-1990,Guerrini-Masini-2001}, provides efficient correctness and sequentialization via local rewriting: it asks that the steps below left result in a single link, connecting both formulas. \defn{Slicing} \cite{Hughes-vanGlabbeek-2005} is a global, geometric criterion: it asks that each \defn{slice}, a choice to remove one subformula of each product along with all connected links, retains a single link.
%
\[
\hspace*{-\parindent}
\hspace*{\fill}
\begin{array}{c@{\qquad\qquad\qquad}c}
	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{A}
		\formula[y=0]{B\*C}
		\Vlink[red]{[-2]1,1}
		\Vlink[red]{[2]1,3}
	\end{tikzpicture}}
~\scoal~
	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{A}
		\formula[y=0]{B\*C}
		\Vlink[red]{1,2}
	\end{tikzpicture}}
\qquad\quad
	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{A}
		\formula[y=0]{B\+C}
		\Vlink[red]{1,1}
	\end{tikzpicture}}
~\scoal~
	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{A}
		\formula[y=0]{B\+C}
		\Vlink[red]{1,2}
	\end{tikzpicture}}
&	
	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{A}
		\begin{scope}[black!40]
		\formula[y=0]{B\*C}
		\Vlink[red]{[-1]1,1}
		\Vlink{[1]1,3}
		\end{scope}
		\formula[y=0]{B{\phantom{{}\*{}}}{\phantom C}}
	\end{tikzpicture}}
\qquad
	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{A}
		\begin{scope}[black!40]
		\formula[y=0]{B\*C}
		\Vlink{[-1]1,1}
		\Vlink[red]{[1]1,3}
		\end{scope}
		\formula[y=0]{{\phantom B}{\phantom{{}\*{}}}C}
	\end{tikzpicture}}
\\
	\text{\small coalescence}
&	\text{\small slicing}
\end{array}	
\hfill
\]
We extend coalescence to both witness nets and unification nets, and slicing to witness nets. Here, we illustrate the former, and leave a discussion of slicing to the conclusion of the paper.

We will distinguish \defn{strict coalescence} ($\scoal$) for witness nets and \defn{unifying coalescence} ($\ucoal$) for unification nets. We first give an example of the former (abbreviating $P(x,y)$ to $Pxy$). In the initial witness net, below left, each link corresponds to a sequent calculus axiom between both linked subformulas, with the substitutions applied.
%
\newcommand\xs{xs}%
\newcommand\xt{xt}%
\newcommand\xy{xy}%
\newcommand\zs{zs}%
\newcommand\zt{zt}%
\begin{center}\vspace{1pt}\(
\hspace*{-\parindent}
\begin{array}{@{}c@{}c@{}c@{}c@{}c@{}}
	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{\forall*{x.}\exists*{y.}{\dual P}*\xy}
		\formula[y=0]{\exists*{z.}P*\zs\*P*\zt}
		\Vlink[red,l,label={$\scriptstyle[x/z,s/y]~~$}]{[-3]3,2}
		\Vlink[blue,r,label={$~~\scriptstyle[x/z,t/y]$}]{[1]3,4}
	\end{tikzpicture}}
& \quad\scoal &
	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{\forall*{x.}\exists*{y.}{\dual P}*\xy}
		\formula[y=0]{\exists*{z.}P*\zs\*P*\zt}
		\Vlink[red,l,label={$\scriptstyle[x/z]~$}]{[-2]2,2}
		\Vlink[blue,r,label={$~~\scriptstyle[x/z,t/y]$}]{[1]3,4}
	\end{tikzpicture}}
& \scoal &
\hfill
	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{\forall*{x.}\exists*{y.}{\dual P}*\xy}
		\formula[y=0]{\exists*{z.}P*\zs\*P*\zt}
		\Vlink[red,l,label={$\scriptstyle[x/z]~$}]{[-2]2,2}
		\Vlink[blue,r,label={$~~\scriptstyle[x/z]$}]{[2]2,4}
	\end{tikzpicture}}
\hfill
\scoal\cdots
\\[6ex]
\text{\scriptsize
	\color{white}
	 \infer{\vphantom{\dual P}}{
	  \color{red}\infer{\seq{\dual P\xs}{P\xs}}{}
       &&
	  \color{blue}\infer{\seq{\dual P\xt}{P\xt}}{}
	 }}
& \hspace*{26pt} &
\text{\scriptsize
	\color{red}\infer[\!\Qrr?s]{\seq{\exists y.\dual P\xy}{P\xs}}{\color{black}\infer{\seq{\dual P\xs}{P\xs}}{}}
~~
	\color{white}\infer{\vphantom{\dual P}}{\color{blue}\infer{\seq{\dual P\xt}{P\xt}}{}}}
& \hspace*{26pt} &
\text{\scriptsize
	\color{red}\infer[\!\Qrr?s]{\seq{\exists y.\dual P\xy}{P\xs}}{\color{black}\infer{\seq{\dual P\xs}{P\xs}}{}}
~~
	\color{blue}\infer[\!\Qrr?t]{\seq{\exists y.\dual P\xy}{P\xt}}{\color{black}\infer{\seq{\dual P\xt}{P\xt}}{}}}
\end{array}
\)\vspace{-3pt}\end{center}
The first two steps (above middle and right) move both links from the subformula $\dual Pxy$ to $\exists y.\dual Pxy$, removing the substitutions $[s/y]$ and $[t/y]$ on $y$, and corresponding to sequent rules $\qrr?s$ and $\qrr?t$. Observe that we maintain the domain of the substitutions on a link as the variables of those existential quantifiers that either linked subformula is (strictly) in scope of.

>>>>

\documentclass[a4paper,UKenglish]{lipics-v2019}

@@

\usepackage{amsmath,amssymb,amsthm}
%\usepackage{latexsym}
\usepackage{microtype}
\usepackage{proof}
%\usepackage{refcount}
%\usepackage{stmaryrd}
%\usepackage{xargs}

@@

\newcommand\defn[1]{\textbf{#1}}

@@

\newcommand\gen{\leq}
\newcommand\coh{\smallfrown}
\newcommand\join{\vee}
%\newcommand\minus{\mathop{\!/\mathchoice{\kern-3pt}{\kern-3pt}{\kern-2.5pt}{\kern-2pt}/\!}}
\newcommand\minus{-}

@@

\author
 {Willem B.\ Heijltjes}
 {University of Bath, United Kingdom\and\url{http://willem.heijltj.es}}
 {} %{w.b.heijltjes@bath.ac.uk}
 {}
 {was supported by EPSRC Project EP/R029121/1 \emph{Typed Lambda-Calculi with Sharing and Unsharing}.}
\author
 {Dominic J.\ D.\ Hughes}
 {Logic Group, UC Berkeley, USA\and\url{http://boole.stanford.edu/~dominic}}
 {} %{dominich@theory.stanford.edu}
 {}
 {} 
\author
 {Lutz Stra\ss burger}
 {Inria Saclay, Palaiseau, France\and LIX, \'Ecole Polytechnique, Palaiseau, France\and \url{http://www.lix.polytechnique.fr/Labo/Lutz.Strassburger}}
 {}
 %{lutz.strassburger@inria.fr}
 {}
 {was supported by the ANR-FWF international grant ANR-5-CE25-0014 \emph{The Fine Structure of Formal Proof Systems and their Computational Interpretations (FISP)}.}

@@

\acknowledgements{We would like to thank the anonymous referees for their constructive feedback. Dominic Hughes thanks his hosts, Wes Holliday and Dana Scott, at the UC Berkeley Logic Group. }

\funding{}

%\section*{Acknowledgements}
%
%We would like to thank the anonymous referees for their constructive feedback. Dominic Hughes thanks his hosts, Wes Holliday and Dana Scott, at the UC Berkeley Logic Group. Willem Heijltjes was supported by EPSRC Project EP/R029121/1 \emph{Typed Lambda-Calculi with Sharing and Unsharing}. Lutz Stra{\ss}burger was supported by the ANR-FWF international
%grant ANR-5-CE25-0014 \emph{The Fine Structure of Formal Proof
%Systems and their Computational Interpretations (FISP)}.

@@

\paragraph*{Additive proof nets}
%
\all\ proof nets \cite[Section 4.10]{Hughes-vanGlabbeek-2005} represent a morphism $A\to B$ by a sequent $\seq{\dual A}B$ plus a \defn{linking}, a relation between the propositional atoms of $\dual A$ (the dual of $A$) and those of $B$. They are \defn{canonical}: they factor out the permutations of sequent calculus, and correspond 1--1 to morphisms of the free category with binary sums and products. \defn{Composition}, of proof nets over $\seq{\dual A}B$ and $\seq{\dual B}C$ to one over $\seq{\dual A}C$, is by the relational composition of their linkings along the dual formulas $B$ and $\dual B$ and captures sequent-calculus cut-elimination. Below are examples of proof nets and their composition.
%
\[
\begin{array}{@{}rc@{\qquad}rc@{}}
	\text{\small diagonal:}
&	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{{\dual a}}
		\formula[y=0]{a\*a}
		\Vlink[red]{[-2]1,1}
		\Vlink[red]{[2]1,3}
	\end{tikzpicture}}
&	\text{\small injection:}	
&   \vc{\begin{tikzpicture}[net]
		\formula[y=1]{{\dual a}}
		\formula[y=0]{a\+b}
		\Vlink[red]{1,1}
	\end{tikzpicture}}
\\ \\
	\text{\small symmetry:}
&	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{{\dual a}\+{\dual b}}
		\formula[y=0]{b\*a}
		\Vlink[red]{1,3}
		\Vlink[red]{3,1}
	\end{tikzpicture}}
&	\text{\small associativity:}
&	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{*{\phantom(}{\dual a}*{\phantom)}
		 {{\+}}*({\dual b}*{\phantom)}
		 {{\+}}*{\phantom(}{\dual c}*)}
		\formula[y=0]{*(a*{\phantom)}
		{{\*}}*{\phantom(}b*)
		{{\*}}*{\phantom(}c*{\phantom)}}
		\Vlink[red]{1,1}
		\Vlink[red]{3,3}
		\Vlink[red]{5,5}
	\end{tikzpicture}}
\end{array}
\qquad
	\text{\small composition:}
	\quad
    \vc{\begin{tikzpicture}[net]
        \formula[y=2.5]{{\dual b}*\+{\dual a}}
        \formula[y=1.5]{a*\**(b*\+b*)}
            \Vlink[red]{1,2}
            \Vlink[red]{2,1}
        \formula[y=1]{{\dual a}*\+*({\dual b}*\*{\dual b}*)}
        \formula[y=0]{b*\*b}
            \Vlink[blue]{[-1]2,[-1]1}
            \Vlink[blue]{[1]2,[-1]2}
            \Vlink[blue]{[-1]3,[1]1}
            \Vlink[blue]{[1]3,[1]2}
    \end{tikzpicture}}
~~\Rightarrow~~
    \vc{\begin{tikzpicture}[net]
        \formula[y=2.5]{{\dual b}*\+{\dual a}}
        \formula[y=0]{b*\*b}
            \Vlink[violet]{[-1]1,1}
            \Vlink[violet]{[1]1,2}
    \end{tikzpicture}}
%\rule[-55pt]{0pt}{110pt}
~~
\]
%
We extend additive proof nets with first-order quantification. Our central challenge is to incorporate the essential content of first-order proof, the \defn{witness assignment} to existential quantifiers. Commonly, as in the sequent calculus rule below left, a witness to $\exists x.B$ is given by an immediate substitution $B[t/x]$. To assign different witnesses in different branches of a proof, the subformula is duplicated first, giving $B[s/x]$ and $B[t/x]$, as below right.
%
\[
	\vc{\infer[\Qrr?t]{\seq A{\exists x.B}}{\seq A{B[t/x]}}}
\hspace{10ex}
\vc{
	\infer[\Qrr*]{\Seq{\exists x.\dual P(x)}{P(s)\*P(t)}}{
	 \infer[\Qrr?s]{\Seq{\exists x.\dual P(x)}{P(s)}}{\infer{\Seq{\dual P(s)}{P(s)}}{}}
	 &&
	 \infer[\Qrr?t]{\Seq{\exists x.\dual P(x)}{P(t)}}{\infer{\Seq{\dual P(t)}{P(t)}}{}}
   }}
\]
%
This is incompatible with a \defn{sequent\,+\,links} proof net design, where the conclusion sequent remains intact, and a subformula $B$ cannot be the subject of substitution or duplication. Instead, we propose two alternative treatments of witnessing terms, embodied in two notions of proof net: \defn{witness nets} and \defn{unification nets}. Our solutions are based on the second author's recent \defn{unification nets} for first-order multiplicative linear logic \cite{Hughes-2018}. Their main feature is to omit existential witnesses altogether, and reconstruct them by unification.
%

@@

\defn{Unification nets} omit any witness information, as illustrated below right. In addition to canonicity, they embody a notion of \defn{generality}: where more than one witness could be assigned, unification nets do not require a definite choice, while witness nets do. Composition is direct, by relational composition. We compare further properties in Figure~\ref{fig:results} in the conclusion, where we also discuss related work and Lambek's notion of \defn{generality} \cite{Lambek-1968-1972}.

\[
	\text{\small a witness net:}
	\quad
	\vc{\begin{tikzpicture}[net]
		\formula[y=1.2]{\exists x.{\dual P}(x)}
		\formula[y=0]{P(s)\*P(t)}
		\Vlink[red,l,label={$\scriptstyle[s/x]~~$}]{[-3]4,1}
		\Vlink[red,r,label={$~\scriptstyle[t/x]$}]{[1]4,6}
	\end{tikzpicture}}
\hspace{11ex}
	\text{\small a unification net:}
	\quad
	\vc{\begin{tikzpicture}[net]
		\formula[y=1.2]{\exists x.{\dual P}(x)}
		\formula[y=0]{P(s)\*P(t)}
		\Vlink[red]{[-3]4,1}
		\Vlink[red]{[1]4,6}
	\end{tikzpicture}}
\]

\paragraph*{Background}

Additive linear logic is combinatorially rich, yet well-behaved and tractable: proof search \cite{Galmiche-Marion-1995} and proof net correctness \cite{Heijltjes-Hughes-2015} for a net over $\seq AB$ are linear in $|A|\times|B|$ (with $|A|$ the size of the syntax tree of $A$). Proof nets remain canonical and equally tractable when extended with the two units \cite{Heijltjes-2011,Heijltjes-Hughes-2015}, and the first-order case is merely NP-complete \cite{Heijltjes-Hughes-2015}.

@@

At the heart of a theory of proof nets is the question of \emph{proof identity}: when are two proofs equivalent? The answer determines which proofs should map onto the same proof net. The introduction of quantifiers creates an interesting issue: if two proofs differ by an immaterial choice of existential witness, should they be equivalent? For example, to prove the sequent $\seq{\exists x.P(x)}{\exists y.\dual P(y)}$ both quantifiers must receive \emph{the same} witness, as in the following two proofs, but \emph{any} witness will do.
\[
	\vc{\infer={\seq{\exists x.P(x)}{\exists y.\dual P(y)}}{\infer{\seq{P(s)}{\dual P(s)}}{}}}
	\quad
	\stackrel?\equiv
	\quad
	\vc{\infer={\seq{\exists x.P(x)}{\exists y.\dual P(y)}}{\infer{\seq{P(t)}{\dual P(t)}}{}}}
\]
The issue is more pronounced where quantifiers are \emph{vacuous}, $\exists x.A$ with $x$ not free in $A$. The proofs below left can only be distinguished even syntactically because the $\exists\mathrm R$ rule makes the instantiating witness explicit. Below right is an interesting intermediate variant: the witness $s$ or $t$ can be observed without explicit annotation in the $\exists\mathrm R$ rule, but the choice is equally immaterial to the content of the proof as when the quantifier were vacuous.
\[
	\vc{\infer[\!\!\scriptstyle{\qrr?s}]{\seq{\exists x.P}{\dual P}}{\infer{\seq P{\dual P}}{}}}
	\quad
	\stackrel?\equiv
	\quad	
	\vc{\infer[\!\!\scriptstyle{\qrr?t}]{\seq{\exists x.P}{\dual P}}{\infer{\seq P{\dual P}}{}}}\qquad
	\vc{
	 \infer[\!\!\scriptstyle{\qrr?s}]{\seq{\exists x.P\+Q(x)}{\dual P}}{
	  \infer[\!\!\scriptstyle{\qrr+1}]{\seq{P\+Q(s)}{\dual P}}{\infer{\seq P{\dual P}}{}}
	}}
	\quad
	\stackrel?\equiv
	\quad	
	\vc{
	 \infer[\!\!\scriptstyle{\qrr?t}]{\seq{\exists x.P\+Q(x)}{\dual P}}{
	  \infer[\!\!\scriptstyle{\qrr+1}]{\seq{P\+Q(t)}{\dual P}}{\infer{\seq P{\dual P}}{}}
	}}
\]
In this paper we will not attempt to settle the question of proof identity. Rather, our two notions of proof net each represent a natural and coherent perspective, at either end of the spectrum. \emph{Witness nets} make all existential witnesses explicit, including those to vacuous quantifiers, rejecting all three equivalences above. \emph{Unification nets} leave all witnesses implicit, thus identifying all proofs modulo witness assignment, and validating all three equivalences.

@@

Additive proof nets have two natural correctness criteria. \defn{Coalescence} \cite{Heijltjes-Hughes-2015,Hughes-Heijltjes-2016}, a counterpart to multiplicative \emph{contractibility} \cite{Danos-1990,Guerrini-Masini-2001}, provides efficient correctness and sequentialization via local rewriting: it asks that the steps below left result in a single link, connecting both formulas. \defn{Slicing} \cite{Hughes-vanGlabbeek-2005} is a global, geometric criterion: it asks that each \defn{slice}, a choice to remove one subformula of each product along with all connected links, retains a single link.
\[
\begin{array}{c@{\qquad\qquad\qquad}c}
	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{A}
		\formula[y=0]{B\*C}
		\Vlink[red]{[-2]1,1}
		\Vlink[red]{[2]1,3}
	\end{tikzpicture}}
~\scoal~
	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{A}
		\formula[y=0]{B\*C}
		\Vlink[red]{1,2}
	\end{tikzpicture}}
\qquad\quad
	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{A}
		\formula[y=0]{B\+C}
		\Vlink[red]{1,1}
	\end{tikzpicture}}
~\scoal~
	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{A}
		\formula[y=0]{B\+C}
		\Vlink[red]{1,2}
	\end{tikzpicture}}
&	
	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{A}
		\begin{scope}[black!40]
		\formula[y=0]{B\*C}
		\Vlink[red]{[-1]1,1}
		\Vlink{[1]1,3}
		\end{scope}
		\formula[y=0]{B{\phantom{{}\*{}}}{\phantom C}}
	\end{tikzpicture}}
\qquad
	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{A}
		\begin{scope}[black!40]
		\formula[y=0]{B\*C}
		\Vlink{[-1]1,1}
		\Vlink[red]{[1]1,3}
		\end{scope}
		\formula[y=0]{{\phantom B}{\phantom{{}\*{}}}C}
	\end{tikzpicture}}
\\
	\text{\small coalescence}
&	\text{\small slicing}
\end{array}
\]
We extend coalescence to both witness nets and unification nets, and slicing to witness nets. Here, we illustrate the former, and leave a discussion of slicing to the conclusion of the paper.

We will distinguish \defn{strict coalescence} ($\scoal$) for witness nets and \defn{unifying coalescence} ($\ucoal$) for unification nets. We first give an example of the former (abbreviating $P(x,y)$ to $Pxy$). In the initial witness net, below left, each link corresponds to a sequent calculus axiom between both linked subformulas, with the substitutions applied.
%
\newcommand\xs{xs}%
\newcommand\xt{xt}%
\newcommand\xy{xy}%
\newcommand\zs{zs}%
\newcommand\zt{zt}%
\[
\begin{array}{@{}c@{}c@{}c@{}c@{}c@{}}
	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{\forall*{x.}\exists*{y.}{\dual P}*\xy}
		\formula[y=0]{\exists*{z.}P*\zs\*P*\zt}
		\Vlink[red,l,label={$\scriptstyle[x/z,s/y]~~$}]{[-3]3,2}
		\Vlink[blue,r,label={$~~\scriptstyle[x/z,t/y]$}]{[1]3,4}
	\end{tikzpicture}}
& \quad\scoal &
	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{\forall*{x.}\exists*{y.}{\dual P}*\xy}
		\formula[y=0]{\exists*{z.}P*\zs\*P*\zt}
		\Vlink[red,l,label={$\scriptstyle[x/z]~$}]{[-2]2,2}
		\Vlink[blue,r,label={$~~\scriptstyle[x/z,t/y]$}]{[1]3,4}
	\end{tikzpicture}}
& \scoal &
\hfill
	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{\forall*{x.}\exists*{y.}{\dual P}*\xy}
		\formula[y=0]{\exists*{z.}P*\zs\*P*\zt}
		\Vlink[red,l,label={$\scriptstyle[x/z]~$}]{[-2]2,2}
		\Vlink[blue,r,label={$~~\scriptstyle[x/z]$}]{[2]2,4}
	\end{tikzpicture}}
\hfill
\scoal\cdots
\\[6ex]
\text{\scriptsize
	\color{white}
	 \infer{\vphantom{\dual P}}{
	  \color{red}\infer{\seq{\dual P\xs}{P\xs}}{}
       &&
	  \color{blue}\infer{\seq{\dual P\xt}{P\xt}}{}
	 }}
& \hspace*{26pt} &
\text{\scriptsize
	\color{red}\infer[\!\Qrr?s]{\seq{\exists y.\dual P\xy}{P\xs}}{\color{black}\infer{\seq{\dual P\xs}{P\xs}}{}}
~~
	\color{white}\infer{\vphantom{\dual P}}{\color{blue}\infer{\seq{\dual P\xt}{P\xt}}{}}}
& \hspace*{26pt} &
\text{\scriptsize
	\color{red}\infer[\!\Qrr?s]{\seq{\exists y.\dual P\xy}{P\xs}}{\color{black}\infer{\seq{\dual P\xs}{P\xs}}{}}
~~
	\color{blue}\infer[\!\Qrr?t]{\seq{\exists y.\dual P\xy}{P\xt}}{\color{black}\infer{\seq{\dual P\xt}{P\xt}}{}}}
\end{array}
\]
The first two steps (above middle and right) move both links from the subformula $\dual Pxy$ to $\exists y.\dual Pxy$, removing the substitutions $[s/y]$ and $[t/y]$ on $y$, and corresponding to sequent rules $\qrr?s$ and $\qrr?t$. Observe that we maintain the domain of the substitutions on a link as the variables of those existential quantifiers that either linked subformula is (strictly) in scope of.
